Paper 2 !DRAFT! Do not circulate!

### 2015-01-01 Articles/Article_EcolEcon_2014/Paper2.Rnw
Clarified how fraction of GDP is calculated.

The term $\mathrm{e}^{\lambda (t-t_0)}$ is known as total factor productivity~($A$),
and $\lambda$ is the Solow residual. 
Assuming that parameters $\theta$, $\alpha$, and $\beta$ are known from parameter estimation,
we can estimate the value of $\lambda$ as follows.
First, assume constant returns to scale such that $\gamma = 1 - \alpha - \beta$.
Then, take the natural logarithm ($\ln$) of Equation~\ref{eq:CD} to obtain
%
\begin{equation} \label{eq:lnCD}
  \ln y = \ln \theta + \ln A + \alpha \ln k + \beta \ln l + (1 - \alpha - \beta) \ln e \;.
\end{equation}
%
By taking the derivative of Equation~\ref{eq:lnCD} with respect to time ($t$) 
and noting that model parameters $\theta$, $\alpha$, and $\beta$ are 
constant with respect to time, we obtain
%
\begin{equation} \label{eq:ddtlnCD}
  \frac{1}{y}\frac{\mathrm{d}y}{\mathrm{d}t} 
      = \frac{1}{A}\frac{\mathrm{d}A}{\mathrm{d}t} 
      + \alpha \frac{1}{k}\frac{\mathrm{d}k}{\mathrm{d}t} 
      + \beta \frac{1}{l}\frac{\mathrm{d}l}{\mathrm{d}t} 
      + (1 - \alpha - \beta) \frac{1}{e}\frac{\mathrm{d}e}{\mathrm{d}t} \;.
\end{equation}
%
Solving for the Solow Residual term gives
%
\begin{equation} \label{eq:solow_resid}
  \frac{1}{A}\frac{\mathrm{d}A}{\mathrm{d}t} 
      = \frac{1}{y}\frac{\mathrm{d}y}{\mathrm{d}t} 
      - \left[
              \alpha \frac{1}{k}\frac{\mathrm{d}k}{\mathrm{d}t} 
              + \beta \frac{1}{l}\frac{\mathrm{d}l}{\mathrm{d}t} 
              + (1 - \alpha - \beta) \frac{1}{e}\frac{\mathrm{d}e}{\mathrm{d}t} 
        \right] \;.
\end{equation}
%
Recognizing that $A \equiv \mathrm{e}^{\lambda (t-t_0)}$ gives
%
\begin{equation}
  \frac{1}{A}\frac{\mathrm{d}A}{\mathrm{d}t} 
    = \frac{1}{\mathrm{e}^{\lambda (t-t_0)}} \; \lambda \mathrm{e}^{\lambda (t-t_0)}
    = \lambda \; ,
\end{equation}
%
which can be substituted into Equation~\ref{eq:solow_resid} to find
%
\begin{equation} \label{eq:lambda}
  \lambda 
    = \frac{1}{y}\frac{\mathrm{d}y}{\mathrm{d}t} 
    - \left[
              \alpha \frac{1}{k}\frac{\mathrm{d}k}{\mathrm{d}t} 
              + \beta \frac{1}{l}\frac{\mathrm{d}l}{\mathrm{d}t} 
              + (1 - \alpha - \beta) \frac{1}{e}\frac{\mathrm{d}e}{\mathrm{d}t} 
      \right] \;.
\end{equation}
%
Equation~\ref{eq:lambda} applies for any instant in time.

If we approximate derivatives in Equation~\ref{eq:lambda} with forward differences
between times $i$ and $j$, we find
%
\begin{equation} \label{eq:lambdaij}
  \lambda_{i,j}
    \approx \frac{1}{y_i}\frac{y_j - y_i}{t_j - t_i} 
    - \left[
            \alpha \frac{1}{k_i}\frac{k_j - k_i}{t_j - t_i} 
            + \beta \frac{1}{l_i}\frac{l_j - l_i}{t_j - t_i} 
            + (1 - \alpha - \beta) \frac{1}{e_i}\frac{e_j - e_i}{t_j - t_i} 
      \right] \;,
\end{equation}
% 
where $\lambda_{i,j}$ approximates the true, instantaneous value of $\lambda$
between times $i$ and $j$.

Interestingly, 
%
\begin{equation}
  \lambda_{1,n}
    \neq \sum_{i=2}^{n} \lambda_{i,i-1}  \;.
\end{equation}
% 
Rather,
%
\begin{equation} \label{eq:lambda1n}
  \lambda_{1,n}
    \approx \frac{1}{y_1}\frac{y_n - y_1}{t_n - t_1} 
      - \left[ 
            \alpha \frac{1}{k_1}\frac{k_n - k_1}{t_n - t_1} 
            + \beta \frac{1}{l_1}\frac{l_n - l_1}{t_n - t_1} 
            + (1 - \alpha - \beta) \frac{1}{e_1}\frac{e_n - e_1}{t_n - t_1} 
      \right] \; ,
\end{equation}
% 
which will become increasingly inaccurate over large time spans,
because $y_1$, $k_1$, $l_1$, and $e_1$ will be less representative
of the average value of $y$, $k$, $l$, and $e$ in the time span, respectively.
This suggests that an averaging approach such as
%
\begin{equation}
  \lambda_{1,n}
    \approx \frac{\sum_{i=2}^{n} \lambda_{i,i-1}}{n-1}
\end{equation}
%
is a better approximation of Equation~\ref{eq:lambda1n},
which is, itself, an approximation of the true value of $\lambda$ 
given by Equation~\ref{eq:lambda}.

@@

<<CD_solowPerc_time_series, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=6, fig.cap='Solow residual time series.'>>=
# Create a time series graph of the solowPerc values
ggplot(data=fdlambda, aes(x=Year, y=solowFrac, color=Source)) + 
  geom_point(size = 1.0) +
  facet_grid(Country ~ energyType) +
  labs(x = "", y = "Fraction of GDP growth explained by Solow residual") + 
  scale_y_continuous(limits=c(-0.5, 2)) +
  xy_theme()
@

>>>>

The term $\mathrm{e}^{\lambda (t-t_0)}$ is known as total factor productivity~($A$),
and $\lambda$ is the Solow residual. 
Assuming that parameters $\theta$, $\alpha$, and $\beta$ are known from parameter estimation,
we can estimate the value of $\lambda$ as follows.
First, assume constant returns to scale such that $\gamma = 1 - \alpha - \beta$.
Then, take the natural logarithm ($\ln$) of Equation~\ref{eq:CDe} to obtain
%
\begin{equation} \label{eq:lnCD}
  \ln y = \ln \theta + \ln A + \alpha \ln k + \beta \ln l + (1 - \alpha - \beta) \ln e \;.
\end{equation}
%
By taking the derivative of Equation~\ref{eq:lnCD} with respect to time ($t$) 
and noting that model parameters $\theta$, $\alpha$, and $\beta$ are 
constant with respect to time, we obtain
%
\begin{equation} \label{eq:ddtlnCD}
  \frac{1}{y}\frac{\mathrm{d}y}{\mathrm{d}t} 
      = \frac{1}{A}\frac{\mathrm{d}A}{\mathrm{d}t} 
      + \alpha \frac{1}{k}\frac{\mathrm{d}k}{\mathrm{d}t} 
      + \beta \frac{1}{l}\frac{\mathrm{d}l}{\mathrm{d}t} 
      + (1 - \alpha - \beta) \frac{1}{e}\frac{\mathrm{d}e}{\mathrm{d}t} \;.
\end{equation}
%
Solving for the Solow Residual term gives
%
\begin{equation} \label{eq:solow_resid}
  \frac{1}{A}\frac{\mathrm{d}A}{\mathrm{d}t} 
      = \frac{1}{y}\frac{\mathrm{d}y}{\mathrm{d}t} 
      - \left[
              \alpha \frac{1}{k}\frac{\mathrm{d}k}{\mathrm{d}t} 
              + \beta \frac{1}{l}\frac{\mathrm{d}l}{\mathrm{d}t} 
              + (1 - \alpha - \beta) \frac{1}{e}\frac{\mathrm{d}e}{\mathrm{d}t} 
        \right] \;.
\end{equation}
%
Recognizing that $A \equiv \mathrm{e}^{\lambda (t-t_0)}$ gives
%
\begin{equation}
  \frac{1}{A}\frac{\mathrm{d}A}{\mathrm{d}t} 
    = \frac{1}{\mathrm{e}^{\lambda (t-t_0)}} \; \lambda \mathrm{e}^{\lambda (t-t_0)}
    = \lambda \; ,
\end{equation}
%
which can be substituted into Equation~\ref{eq:solow_resid} to find
%
\begin{equation} \label{eq:lambda}
  \lambda 
    = \frac{1}{y}\frac{\mathrm{d}y}{\mathrm{d}t} 
    - \left[
              \alpha \frac{1}{k}\frac{\mathrm{d}k}{\mathrm{d}t} 
              + \beta \frac{1}{l}\frac{\mathrm{d}l}{\mathrm{d}t} 
              + (1 - \alpha - \beta) \frac{1}{e}\frac{\mathrm{d}e}{\mathrm{d}t} 
      \right] \;.
\end{equation}
%
Equation~\ref{eq:lambda} applies for any instant in time.

The fraction of GDP growth explained by the Solow residual~($f$)
can be given as
%
\begin{equation} \label{eq:frac_growth_explained_by_solow}
  f \equiv \frac{\lambda}{\frac{1}{y}\frac{\mathrm{d}y}{\mathrm{d}t}} 
            = 1 - \frac{\alpha \frac{1}{k}\frac{\mathrm{d}k}{\mathrm{d}t} 
                      + \beta \frac{1}{l}\frac{\mathrm{d}l}{\mathrm{d}t} 
                      + (1 - \alpha - \beta) \frac{1}{e}\frac{\mathrm{d}e}{\mathrm{d}t}}
                  {\frac{\lambda}{\frac{1}{y}\frac{\mathrm{d}y}{\mathrm{d}t}}}
\end{equation}
%
If we approximate derivatives in Equation~\ref{eq:lambda} with forward differences
between times $i$ and $j$, we find
%
\begin{equation} \label{eq:lambdaij}
  \lambda_{i,j}
    = \frac{1}{y_i}\frac{y_j - y_i}{t_j - t_i} 
    - \left[
            \alpha \frac{1}{k_i}\frac{k_j - k_i}{t_j - t_i} 
            + \beta \frac{1}{l_i}\frac{l_j - l_i}{t_j - t_i} 
            + (1 - \alpha - \beta) \frac{1}{e_i}\frac{e_j - e_i}{t_j - t_i} 
      \right] \;,
\end{equation}
% 
where $\lambda_{i,j}$ approximates the true, instantaneous value of $\lambda$
between times $i$ and $j$.

Interestingly, 
%
\begin{equation}
  \lambda_{1,n}
    \neq \sum_{i=2}^{n} \lambda_{i,i-1}  \;.
\end{equation}
% 
Rather,
%
\begin{equation} \label{eq:lambda1n}
  \lambda_{1,n}
    = \frac{1}{y_1}\frac{y_n - y_1}{t_n - t_1} 
      - \left[ 
            \alpha \frac{1}{k_1}\frac{k_n - k_1}{t_n - t_1} 
            + \beta \frac{1}{l_1}\frac{l_n - l_1}{t_n - t_1} 
            + (1 - \alpha - \beta) \frac{1}{e_1}\frac{e_n - e_1}{t_n - t_1} 
      \right] \; ,
\end{equation}
% 
which will become increasingly inaccurate over large time spans,
because $y_1$, $k_1$, $l_1$, and $e_1$ will be less representative
of the average value of $y$, $k$, $l$, and $e$ in the time span, respectively.
This suggests that an averaging approach such as
%
\begin{equation}
  \lambda_{1,n}
    = \frac{\sum_{i=2}^{n} \lambda_{i,i-1}}{n-1}
\end{equation}
%
is a better approximation of Equation~\ref{eq:lambda1n},
which is, itself, an approximation of the true value of $\lambda$ 
given by Equation~\ref{eq:lambda}.

Using forward differences to estimate $f$ yields
%
\begin{equation} \label{eq:f_finite_diff}
  f_{i,j} = \frac{\lambda_{i,j}}{\frac{1}{y_i}\frac{y_j - y_i}{t_j - t_i}}
       = 1 - \frac{\alpha \frac{1}{k_i}\frac{k_j - k_i}{t_j - t_i} 
            + \beta \frac{1}{l_i}\frac{l_j - l_i}{t_j - t_i} 
            + (1 - \alpha - \beta) \frac{1}{e_i}\frac{e_j - e_i}{t_j - t_i} }
            {\frac{1}{y_i}\frac{y_j - y_i}{t_j - t_i}}
\end{equation}
%
which is an estimate to the true value of $f$ given by Equation~\ref{eq:frac_growth_explained_by_solow}.

@@

<<CD_solowPerc_time_series, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=6, fig.cap='Fraction of GDP growth explained by Solow residual~($\\lambda$).'>>=
# Create a time series graph of the solowPerc values
ggplot(data=fdlambda, aes(x=Year, y=solowFrac, color=Source)) + 
  geom_point(size = 1.0) +
  facet_grid(Country ~ energyType) +
  labs(x = "", y = "Fraction of GDP growth explained by Solow residual") + 
  scale_y_continuous(limits=c(-0.5, 2)) +
  xy_theme()
@
